% =============================================================================
% subselect — methodology log
%
% This file records finalized methodology decisions for the subselect framework.
% It is the source-of-truth that the EGU 2026 poster and the follow-up paper
% will both quote. See ../docs/documentation_workflow.md for what goes here
% and what doesn't.
%
% Compile: pdflatex methods.tex
% =============================================================================

\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{geometry}
\geometry{margin=2.2cm}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[round]{natbib}
\usepackage{enumitem}
\setlist{itemsep=2pt, topsep=4pt}

\title{\texttt{subselect} --- Methodology Log}
\author{Athanasios Tsilimigkras \\
        Technical University of Crete, Chania, Greece}
\date{Last updated: \today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This document records the finalized methodology decisions of the \texttt{subselect}
framework for country-scale CMIP6 sub-ensemble selection. Each decision lists the
question, the alternatives considered, the chosen option, the rationale, the
defining equation or pseudocode, the implementation pointer, and the date.
The document is appended to as decisions are made, never edited retroactively.
It feeds the EGU~2026 poster (abstract EGU26-14502)
and the follow-up framework paper.
\end{abstract}

\tableofcontents

% =============================================================================
\section{Historical performance}\label{sec:performance}
% =============================================================================

The historical-performance dimension is reused unchanged from the published Greece
paper. See \texttt{../docs/historical\_performance.md} for the complete specification
of the variables, reference dataset, evaluation window, and aggregation choices.

% --- entries appended below as performance methodology evolves -----------------

\subsection{Data sources and regridding}\label{sec:performance-data-sources}

\paragraph{Question.} Which CMIP6 grid and which observation product back
the published Greece paper's HPS values, and what regridding choices stand
behind them?

\paragraph{Alternatives considered.}
\begin{itemize}
  \item Among the precomputed CMIP6 products on disk (\texttt{monthly},
        \texttt{monthly\_new}, \texttt{monthly\_original}; pre-restructure
        names): each was a separate ingest with its own time-conventions and
        decoding choices.
  \item Among the observation products: raw ISIMIP3a
        (\texttt{ISIMIP3a/monthly/}, native $\sim$0.5\textdegree{} grid),
        single-grid upscaled (\texttt{ISIMIP3a/monthly\_upscaled/}, common
        grid for all models), per-CMIP6-model upscaled
        (\texttt{ISIMIP3a/monthly\_cmip6\_upscaled/<MODEL>/}, model-by-model
        grid).
  \item Among regridding schemes: \texttt{xesmf} bilinear, \texttt{xesmf}
        conservative, scipy nearest-neighbor, \texttt{cdo} remapcon /
        remapnn / remapbil.
\end{itemize}

\paragraph{Choice.} CMIP6 from \texttt{Data/CMIP6/monthly/} (post-restructure
of \texttt{monthly\_new/}) on each model's native grid; observations from
\texttt{Data/reference/monthly\_cmip6\_upscaled/<MODEL>/} for the per-pixel
model-vs-obs metrics; observations regridded with a per-variable scheme
that is \emph{conservative interpolation for \texttt{pr}} and
\emph{nearest-neighbor for \texttt{tas} / \texttt{psl} / \texttt{tasmax}}
(per the inline comment in \texttt{legacy/cmip6-greece/GR\_model\_performance\_HM.ipynb}
line 55). The 1995--2014 evaluation window per \texttt{Config.eval\_window}
is sliced \emph{after} loading.

\paragraph{Rationale.} The choices are inherited from the published Greece
paper~\citep{tsilimigkras2026subselection} and verified by the regression
test, not re-litigated. The pipeline does \emph{no} on-the-fly regridding;
regridding is upstream pre-processing performed during the ISIMIP3a
ingest, and \texttt{subselect} consumes the artefacts as-is. The tooling
that produced the upscaled files is not part of the framework's
dependency surface.

\paragraph{Implementation.} \texttt{subselect.io.load\_cmip6} reads
\texttt{Data/CMIP6/monthly/<var>/<scenario>/<MODEL>\_*.nc};
\texttt{subselect.io.load\_w5e5} reads \texttt{Data/reference/monthly\_cmip6\_upscaled/<MODEL>/<var>\_*\_<MODEL>.nc}
via \texttt{REFERENCE\_FILENAME\_TEMPLATES}. Period slicing happens
inside \texttt{subselect.performance.compute\_hps}.

\paragraph{Decided.} 2026-05-01, in version 0.1.0.

\subsection{Observation reference grid for $\sigma_{\text{obs}}$ vs per-pixel metrics}\label{sec:performance-sigma-obs-grid}

\paragraph{Question.} Which observation reference grid is used for the
$\sigma_{\text{obs}}$ scalar in the TSS denominator, versus the per-pixel
bias / RMSE / correlation / bias\_score metrics that compare a model to
observations grid-cell-by-grid-cell?

\paragraph{Alternatives considered.}
\begin{itemize}
  \item \textbf{Both metrics from the per-CMIP6-model upscaled product.} One
        $\sigma_{\text{obs}}$ value per (variable, season, \emph{model}),
        regridded variance bakes into TSS.
  \item \textbf{Both metrics from a single common grid.} One
        $\sigma_{\text{obs}}$ per (variable, season), but per-pixel
        comparisons are no longer on each model's native grid --- model
        fields would need to be regridded to the common grid, which the
        framework explicitly avoids.
  \item \textbf{Two-tier reference (chosen).} $\sigma_{\text{obs}}$ from a
        single common grid; per-pixel metrics from each model's own grid.
\end{itemize}

\paragraph{Choice.} Two-tier reference: $\sigma_{\text{obs}}$ from the
single-grid upscaled product
(\texttt{Data/reference/monthly\_upscaled/<var>\_*\_cmip6\_upscaled.nc}),
per-pixel metrics from the per-CMIP6-model upscaled product
(\texttt{Data/reference/monthly\_cmip6\_upscaled/<MODEL>/<var>\_*\_<MODEL>.nc}).

\paragraph{Rationale.} Decouples TSS values from model-specific regridding
variance. Per-model $\sigma_{\text{obs}}$ would couple TSS to the
idiosyncrasies of each model's grid (cell sizes, $\cos\varphi$ mass
distribution, regridding artefacts at coastlines), so the cross-model
min--max normalisation that feeds HMperf would be measuring a mixture of
\emph{model fidelity} and \emph{regridding fidelity}. A single common
$\sigma_{\text{obs}}$ removes that confound and preserves cross-model
comparability for the downstream normalisation. The per-pixel comparison
metrics (bias, RMSE, correlation, bias\_score) must stay on each model's
own grid because they compare \emph{model field} to \emph{obs field} cell
by cell, and the framework does not regrid models. The two-product setup
costs the storage of one extra reference dataset on disk and a second
loader path in \texttt{subselect.io}.

\paragraph{Initial parity caveat (corrected in 0.1.0).} The published
Greece pipeline reduced the single-grid per-pixel std-over-months map
to $\sigma_{\text{obs}}$ via an \emph{unweighted}
\texttt{.mean(dim=['lat', 'lon'])}, inconsistent with the
$\cos\varphi$-weighted spatial mean used everywhere else. Version 0.1.0
restored the $\cos\varphi$ weighting; see \S\ Methodology corrections
(post-paper).

\paragraph{Implementation.}
\texttt{subselect.io.load\_single\_grid\_w5e5} (one file per variable,
common grid) for $\sigma_{\text{obs}}$;
\texttt{subselect.io.load\_w5e5} (per-model upscaled) for the per-pixel
comparison. \texttt{Config.single\_grid\_reference\_root} (default
\texttt{<data\_root>/reference/monthly\_upscaled}) and
\texttt{Config.reference\_root} (default
\texttt{<data\_root>/reference/monthly\_cmip6\_upscaled}) keep the two
products separately addressable; both are TOML-overridable.

\paragraph{Decided.} 2026-05-01, in version 0.1.0. Both initial parity
caveats (the unweighted spatial mean and the cmip6-grid upscaled
$\sigma_{\text{obs}}$ source) were corrected in version 0.1.0; see
\S\ Methodology corrections (post-paper).

\subsection{Bias-map observed-mean panel: native 0.5\textdegree{} W5E5}\label{sec:performance-bias-map-obs}

The bias-map figure
(\texttt{subselect.viz.performance\_figs.fig\_bias\_maps\_per\_variable})
has a top-row \emph{observed-mean} panel that displays the W5E5
reference field at high resolution for visual context, separate from
the per-model bias panels below it. The observed-mean panel is read
from the native 0.5\textdegree{} W5E5 product (360~$\times$~720), not
from the cmip6-grid upscaled product used for $\sigma_{\text{obs}}$.

The bias-map adapter
(\texttt{subselect.viz.\_data\_adapters.build\_bias\_maps\_state})
populates \texttt{observed\_maps[period][variable]} via
\texttt{subselect.io.load\_native\_w5e5}, which resolves to
\texttt{<data\_root>/reference/monthly\_05/<var>\_*.nc}. This yields
the high-resolution observed-mean panel that pairs naturally with the
per-model bias maps (\textasciitilde 520 native pixels in the Greece
bbox, vs the \textasciitilde 49 pixels the cmip6-grid upscaled product
would have produced). Per-model bias panels remain unchanged --- they
use the per-CMIP6-model upscaled obs (\texttt{load\_w5e5}) since the
bias is computed cell-by-cell on each model's own grid.

\paragraph{Implementation.}
\texttt{subselect.io.load\_native\_w5e5(variable)} returns the native
0.5\textdegree{} dataset; the loader sorts by lat/lon on read because
ISIMIP3a files are stored with descending latitude. Decided 2026-05-01,
in version 0.1.0.

\subsection{Methodology corrections (post-paper)}\label{sec:performance-corrections}

The published Greece pipeline carried two minor inconsistencies in the
$\sigma_{\text{obs}}$ scalar; version 0.1.0 corrected both. The
corrections shift TSS values slightly (typically $<\!1\%$ per model)
and therefore shift HMperf and the model rankings. The corrected
behaviour is canonical from version 0.1.0 onward. Reference outputs
from the published Greece pipeline are archived under
\texttt{documentation/historical\_paper\_outputs/} for traceability.

\paragraph{Correction 1: $\cos\varphi$-weighted spatial mean.}
The published pipeline reduced the per-pixel std-over-months map to
the $\sigma_{\text{obs}}$ scalar via an \emph{unweighted}
\texttt{.mean(dim=['lat', 'lon'])}, inconsistent with the
$\cos\varphi$-weighted spatial mean used everywhere else in the
pipeline (per-model \texttt{std\_dev}, \texttt{bias\_score}, all
per-pixel comparison metrics). The corrected pipeline routes the
$\sigma_{\text{obs}}$ spatial mean through
\texttt{subselect.performance.\_spatial\_weighted\_mean}, the same
$\cos\varphi$-weighted helper that every other spatial mean uses.

\paragraph{Correction 2: native 0.5\textdegree{} reference for $\sigma_{\text{obs}}$.}
The published pipeline read $\sigma_{\text{obs}}$ from the cmip6-grid
upscaled product
(\texttt{Data/reference/monthly\_upscaled/<var>\_*\_cmip6\_upscaled.nc},
1.25\textdegree~$\times$~1.875\textdegree, 144~$\times$~192). The
corrected pipeline reads from the native 0.5\textdegree{} single-grid
product (\texttt{Data/reference/monthly\_05/<var>\_*.nc},
0.5\textdegree~$\times$~0.5\textdegree, 360~$\times$~720) via
\texttt{subselect.io.load\_native\_w5e5}. The native product is the
direct W5E5 obs without an upscaling pre-processing step, so
$\sigma_{\text{obs}}$ no longer carries upscaling-imprint variance.

\paragraph{Why these are correct.}
Both corrections remove confounds that should not be in the
$\sigma_{\text{obs}}$ scalar. (1) An unweighted spatial mean
double-counts high-latitude pixels relative to their actual area; the
rest of the pipeline uses $\cos\varphi$ weights for exactly this reason,
and the original inconsistency was an oversight rather than a
methodology choice. (2) Computing $\sigma_{\text{obs}}$ on a coarser
grid than the per-pixel metrics introduces resolution-dependent
smoothing into the TSS denominator that has no physical basis.

\paragraph{Regression contract.}
Version 0.1.0 pins parquet snapshots of the corrected pipeline outputs
(\texttt{compute\_metrics} per variable, \texttt{compute\_hps},
\texttt{compute\_change\_signals}) under
\texttt{tests/fixtures/regression\_corrected/}.
\texttt{tests/test\_regression\_corrected.py} asserts bit-identity at
\texttt{1e-12} against these snapshots.

\paragraph{Decided.} 2026-05-02, in version 0.1.0.

% =============================================================================
\section{Future-response spread}\label{sec:spread}
% =============================================================================

The future-spread dimension is partially inherited from the published Greece
paper~\citep{tsilimigkras2026subselection} (change variables, scenario
default) and partially extended for the framework (formal sub-ensemble
coverage metric).

% --- entries appended below as spread methodology is finalized ----------------

\subsection{Change-signals regression-test tolerance}\label{sec:spread-tolerance}

\paragraph{Question.} What tolerance does the change-signals regression test
use against the published xlsx, and what failure mode justifies a
non-uniform tolerance across the 35-model ensemble?

\paragraph{Choice.} ``ATOL\_TARGET = 1e-6'' for 34/35 models;
``ATOL\_CHANGE\_SIGNALS = 5e-4'' as the test-wide tolerance to absorb a
single-model float32 drift on IPSL-CM6A-LR's psl columns specifically.

\paragraph{Rationale.} CMIP6 NetCDF files store fields as float32, which
are read and promoted to float64 during xarray operations. For 34 models
the promotion order matches the published Greece pipeline exactly, so the
change-signal scalars round-trip bit-identically
(observed max absolute diff $< 10^{-12}$). IPSL-CM6A-LR is the lone
outlier: it uses a noleap calendar, and xarray's time-decoding path for
non-standard calendars promotes float32 $\to$ float64 at a slightly later
operation than for standard calendars. The arithmetic is identical;
only the order of precision promotion differs. Catastrophic cancellation
between two $\sim$$10^5$ Pa pre-industrial and end-of-century pressure
values amplifies the float32 ULP ($\sim$$0.012$ Pa) into the observed
$\sim$$10^{-4}$ relative error on the change signal. tas / tasmax for IPSL
also drift but stay within $2\!\times\!10^{-6}$ because cancellation is
mild for $\sim$$6$ K change signals on $\sim$$300$ K absolute values. The
$5\times 10^{-4}$ tolerance leaves a $\sim$$5\times$ margin past the
observed maximum breach ($1.83 \times 10^{-4}$ on
\texttt{psl\_DJF}); the spread-quadrant labels (cross-model medians on
$\Delta$tas / $\Delta$pr) are unaffected because the float32 noise is
orders of magnitude smaller than the inter-model spread on which the
discrete labels rest.

\paragraph{Implementation.}
\texttt{tests/test\_regression\_greece.py::test\_greece\_change\_signals\_regression}
(uses \texttt{ATOL\_CHANGE\_SIGNALS}). The HPS regression test stays at
\texttt{ATOL\_TARGET}.

\paragraph{Decided.} 2026-05-01, in version 0.1.0.

% =============================================================================
\section{Model independence}\label{sec:independence}
% =============================================================================

The independence dimension is new in the framework; it was not part of the
published Greece paper~\citep{tsilimigkras2026subselection}. Two methods are
scoped: feature-space clustering on a regional climate-response embedding,
and pairwise-RMSE genealogy clustering.

% --- entries appended below as independence methodology is finalized ---------

% =============================================================================
\section{Multi-objective optimization}\label{sec:optimization}
% =============================================================================

The selection engine combines the three dimensions into a scalar cost
(default) and a Pareto front (diagnostic). See \texttt{subselect.optimize}
for the implementation.

% --- entries appended below as optimization methodology is finalized ---------

% =============================================================================
\section{Country cropping}\label{sec:cropping}
% =============================================================================

The framework supports both bounding-box and shapefile-based cropping of
the input fields. See \texttt{subselect.geom} for the implementation.

% --- entries appended below as cropping policy is finalized ------------------

\subsection{Framework default crop rule}\label{sec:cropping-default}

\paragraph{Question.} Which cropping rule should the framework apply by default
when extracting a CMIP6 model's data over a user-specified country?

\paragraph{Alternatives considered.}
\begin{itemize}
  \item \textbf{\texttt{bbox}}: rectangular slice from
        \texttt{country\_codes.json}, no shapefile dependency. The published
        Greece paper used this; it over-includes ocean and neighbouring land,
        especially for archipelagic countries.
  \item \textbf{\texttt{shapefile\_strict}}: binary mask via
        \texttt{rasterio.features.geometry\_mask} with
        \texttt{all\_touched=False} --- a pixel is included only if its
        centroid lies inside the country polygon. Risk: leaves zero pixels
        for small countries on coarse grids (e.g.\ Cyprus on CanESM5
        $\sim$2.8\,\textdegree).
  \item \textbf{\texttt{shapefile\_lenient}}: binary mask via
        \texttt{rasterio.features.geometry\_mask} with
        \texttt{all\_touched=True} --- any pixel touched by the country
        polygon is included. NaN outside the country.
  \item \textbf{\texttt{shapefile\_fractional}}: per-pixel area-fraction
        inside the country polygon, in $[0, 1]$, via per-cell
        \texttt{shapely} intersection. Returns the data un-masked plus a
        weight array. Boundary-precision option; opt-in.
\end{itemize}

\paragraph{Choice.} \texttt{shapefile\_lenient}.

\paragraph{Rationale.} Three reasons. First, \emph{operational simplicity}:
the cropped result is a rectangular \texttt{DataArray} with NaN outside the
country, which xarray and matplotlib both handle natively; the existing
$\cos(\varphi)$ area-weighted spatial-mean code works unchanged because the
\texttt{xr.DataArray.weighted(\dots).mean(\dots)} primitive already ignores
NaN entries. Second, \emph{prototype consistency}: the legacy
\texttt{climpact/shp\_extraction.ipynb} prototype already uses this rule, and
adopting it as the default avoids re-litigating an empirical choice that was
satisfactory in earlier exploratory work. Third, \emph{empirical validation}:
visual comparison of all four methods over Greece, Cyprus, and Switzerland
on CanESM5 and MPI-ESM1-2-HR confirmed that \texttt{shapefile\_lenient}
captures the country footprint sensibly across archipelagic (Greece),
small-island (Cyprus), and mountainous-landlocked (Switzerland) geometries
on coarse and fine native grids. The regression test continues to pin
\texttt{bbox} for parity with the published Greece numbers;
\texttt{shapefile\_fractional} remains available for boundary-precision
use cases that justify the larger interface change to weight-aware metrics.

\paragraph{Definition.} For each pixel $p$ on the model's native grid with
spatial extent $A_p$:
%
\begin{equation}
  m(p) =
  \begin{cases}
    1 & \text{if } A_p \cap \Omega \neq \emptyset, \\
    0 & \text{otherwise},
  \end{cases}
\end{equation}
%
where $\Omega$ is the country polygon from GADM 4.1 (layer \texttt{ADM\_0}).
The mask is computed against the full input grid; both the data and the mask
are then cropped to a mask-aware bounding box (the smallest lat/lon box
enclosing all $m(p)=1$ pixels, padded by one full cell on each side). The
cropped data is masked with NaN where $m(p) = 0$:
%
\begin{equation}
  d_{\text{crop}}(p) =
  \begin{cases}
    d(p) & \text{if } m(p) = 1, \\
    \text{NaN} & \text{if } m(p) = 0.
  \end{cases}
\end{equation}

\paragraph{Implementation.} \texttt{subselect.geom.crop} with
\texttt{method="shapefile\_lenient"} (the default kwarg value); pinned by
\texttt{tests/test\_geom.py::test\_lenient\_includes\_cell\_when\_polygon\_lands\_inside\_a\_single\_cell}
and the overshoot-half-cell regression case alongside it.

\paragraph{Decided.} 2026-04-30, in version 0.1.0.

% =============================================================================
\section{Caching and reproducibility}\label{sec:caching}
% =============================================================================

The framework caches per-country results in two layers: parquet for tabular
metric tables and zarr for fields and matrices, indexed by a small JSON
catalog per scope.

% --- entries appended below as caching/reproducibility decisions are made ----

\subsection{Cache stack: parquet + zarr + SQLite}\label{sec:caching-stack}

\paragraph{Question.} Which storage stack should back the per-country and
ensemble-wide derived artefacts (HPS / BVS tables, change signals, climatology
fields, GWL crossing years, distance matrices)?

\paragraph{Alternatives considered.}
\begin{itemize}
  \item \textbf{Parquet (tabular) + Zarr (arrays) + SQLite (index).}
        Three layers, each with one responsibility. Parquet is language-portable,
        schema-explicit, snappy-compressed; Zarr v2 directory stores serve
        trivially as static files; SQLite is in the Python stdlib.
  \item \textbf{DuckDB on Parquet.} Same parquet files, DuckDB replaces SQLite
        as the catalog \emph{and} runs analytical queries directly against
        the parquet files. Extra dependency.
  \item \textbf{Intake catalog} (\texttt{intake-xarray}). YAML-based,
        declarative, climate-community standard for indexing xarray datasets.
  \item \textbf{HDF5 hierarchies} (\texttt{h5py} or \texttt{netCDF4}).
        Single-file, hierarchical container.
\end{itemize}

\paragraph{Choice.} Parquet (tabular) + Zarr (arrays) + SQLite (index).

\paragraph{Rationale.} Four reasons. First, \emph{workload fit}: tabular
outputs are KB to MB-scale per artefact (the canonical Greece directory is
$\sim$138\,MB across 28 spreadsheets), and the hot path is per-country
DataFrame loads in the Phase~2 cost-function inner loop. DuckDB's value is
analytical SQL across many parquet files; that workload does not exist
here, so SQLite as a pure path-catalog is the right weight class. Second,
\emph{consumer set}: the cache is read only by \texttt{subselect} itself and
the future static web app; \texttt{intake-xarray}'s value is sharing a
declarative catalog with third-party climate tooling, which is not a
requirement. Third, \emph{web-app delivery}: zarr~v2 directory stores serve
trivially over HTTP as static files (the Phase~4 web-app delivery model);
HDF5 single-files require a server runtime to slice. Fourth,
\emph{reproducibility hooks}: cache invalidation by code version is a
single JSON-catalog field; harder to retrofit into intake YAML or HDF5
attributes. Earlier exploratory pickle artefacts were language-locked
and version-fragile; parquet replaces them and removes that dependency.

\paragraph{Definition.} The cache is per-country and indexed by a small
JSON catalog at \texttt{cache/}$\langle c \rangle$\texttt{/catalog.json}.
Each artefact records its on-disk path, the maximum mtime of its declared
inputs at write time, and the time of computation. Layout:
%
\begin{align*}
  \text{tabular}     \quad &\text{cache/}\langle c \rangle / \texttt{parquet/} \langle k \rangle \texttt{.parquet} \\
  \text{dict-tabular} \quad &\text{cache/}\langle c \rangle / \texttt{parquet/} \langle k \rangle / \langle s \rangle \texttt{.parquet} \\
  \text{dataclass}   \quad &\text{cache/}\langle c \rangle / \texttt{parquet/} \langle k \rangle / \{\langle f \rangle \texttt{.parquet}, \texttt{scalars.json}\} \\
  \text{fields}      \quad &\text{cache/}\langle c \rangle / \texttt{zarr/} \langle k \rangle \texttt{.zarr}
\end{align*}
%
where $c$ = country, $k$ = artefact key, $s$ = sub-key, $f$ = dataclass
field. On read, an artefact is invalidated whenever any declared input
file has an mtime newer than the recorded maximum.

\paragraph{Implementation.} \texttt{subselect.cache.Cache} exposes
\texttt{has}, \texttt{is\_fresh}, \texttt{save}, \texttt{load},
\texttt{invalidate}, \texttt{clear}. Artefact kinds are inferred from
value type (\texttt{DataFrame}, \texttt{Series}, \texttt{Dataset},
\texttt{DataArray}, \texttt{dict[str, DataFrame]}, dataclass).
\texttt{zarr<3} pinned in \texttt{environment.yml} and
\texttt{pyproject.toml} for v2 directory-store stability.

\paragraph{Decided.} 2026-05-01, in version 0.1.0. An earlier draft of
the cache layer carried a SQLite catalog; the simpler scope (one
country, one cache, one catalog) made cross-country indexing unnecessary
and the per-scope JSON catalog used in version 0.1.0 onward removes
SQLite's I/O and schema overhead.

\subsection*{Cache scope: global vs per-country}

\paragraph{Question.} Where do per-(model, variable) climatology fields,
$\sigma_\text{obs}$ grids, and per-(model, variable, scenario)
annual-mean fields live?

\paragraph{Choice.} Country-independent artefacts move to
\texttt{cache/\_global/}; country-specific reductions stay under
\texttt{cache/}$\langle c \rangle$\texttt{/}. The \emph{compute-once-globally,
reuse-per-country} convention.

\paragraph{Rationale.} A per-(model, variable) monthly climatology over
the 1995--2014 evaluation window depends only on (model, variable), not
on the country chosen for analysis. The same is true of the corresponding
upscaled-observation climatology, the per-period interannual $\sigma$ map, the
end-of-century and pre-industrial monthly climatologies, and the
per-(model, variable, scenario) annual-mean field 1850--2100. Country
analysis cropping is a slice of the cached field, not a recomputation of
it. Caching at country scope would re-pay the per-model climatology and
annual-mean cost on every new country; caching at global scope pays it
once. After the first country run, every subsequent country reuses the
cached fields.

\paragraph{Layout.}
\begin{align*}
  \text{global} \quad &\text{cache/\_global/}\{ \text{catalog.json},
  \text{zarr/}\langle a \rangle\texttt{.zarr}, \text{parquet/}\langle k \rangle\texttt{.parquet} \} \\
  \text{per-country} \quad &\text{cache/}\langle c \rangle / \{ \text{catalog.json},
  \text{parquet/}, \text{zarr/} \}
\end{align*}

\paragraph{Global artefacts.}
\begin{itemize}
\item Per-(model, variable) historical climatology
  (\texttt{historical\_clim\_\_<var>\_\_<model>.zarr}); 12 months on the
  model's native grid over 1995--2014.
\item Per-(model, variable) upscaled-observation climatology
  (\texttt{obs\_clim\_\_<var>\_\_<model>.zarr}).
\item Per-(model, variable, period) interannual $\sigma$ map
  (\texttt{sigma\_ref\_\_<var>\_\_<model>.zarr}).
\item Per-(model, variable, scenario) end-of-century climatology
  (\texttt{eoc\_clim\_\_<var>\_\_<model>\_\_<ssp>.zarr}); 12 months over 2081--2100.
\item Per-(model, variable) pre-industrial climatology
  (\texttt{pi\_clim\_\_<var>\_\_<model>.zarr}); 12 months over 1850--1899.
\item Per-(model, variable, scenario) annual-mean field
  (\texttt{annual\_field\_\_<var>\_\_<model>\_\_<ssp>.zarr}); 251 years
  on the model's native grid.
\item Native $0.5^\circ$ W5E5 climatology + per-period $\sigma$ field per variable
  (\texttt{native\_sigma\_obs\_\_<var>.zarr}); feeds the $\sigma_\text{obs}$
  scalar in the TSS denominator.
\end{itemize}

\paragraph{Per-country artefacts} are country-mean reductions of the
above: HPS / TSS / BVS metric tables, country annual time-series,
change\_signals, country $\sigma_\text{obs}$ scalars per period, country
profile signals, country-mean monthly climatologies, country-mean bias
fields, and the warming-level / future-anomaly tables.

\paragraph{Implementation.} \texttt{subselect.compute\_global.compute\_global()}
populates the global cache; \texttt{subselect.compute.compute(country)}
auto-builds the global cache on first call and consumes it thereafter.
The \texttt{Cache} class' \texttt{global\_cache} classmethod wraps the
global scope. The CLI exposes \texttt{python -m subselect --global-only}
for explicit warmup and \texttt{--force \{all,country,global\}} for
scope-targeted recompute.

\paragraph{Numerical artefact of the cache architecture.} Compared to
an earlier development draft that performed \emph{load $\rightarrow$
crop $\rightarrow$ time-slice $\rightarrow$ climatology $\rightarrow$
spatial-mean} per-country, the global-cache architecture used from
version 0.1.0 onward computes the climatology globally and then crops:
\emph{climatology (global) $\rightarrow$ crop $\rightarrow$
spatial-mean}. The two operation orders are mathematically equivalent
on a continuous-valued field; on float32 reductions the spread-pipeline
outputs (\texttt{change\_signals}, \texttt{long\_term\_spread},
\texttt{pre\_industrial\_spread}) drift by $\sim$\,$10^{-5}$ absolute /
$10^{-7}$ relative — six significant digits, six orders of magnitude
below any climate-relevant threshold. This is a numerical artefact of
the cache architecture, not a methodology change. The post-cache
values are the canonical regression-test contract.

\paragraph{Cross-scope invalidation.} Per-country artefacts that derive
from global-cache contents include
\texttt{cache/\_global/catalog.json} in their dependency list. Saves
into the global cache rewrite this catalog (its mtime advances), so
the next \texttt{is\_fresh} check on any downstream per-country
artefact returns \texttt{False} and triggers a recompute. This
prevents the failure mode where a global-cache implementation change
or partial rebuild leaves stale per-country derivations: any global
mutation now propagates invalidation downstream. The rule applies to
\texttt{observed\_std\_dev}, \texttt{performance\_metrics\_<var>},
\texttt{monthly\_means\_<var>\_*}, \texttt{composite\_hps},
\texttt{composite\_hps\_full}, \texttt{change\_signals},
\texttt{long\_term\_spread}, \texttt{pre\_industrial\_spread}, and
\texttt{annual\_timeseries\_<var>}. A unit test in
\texttt{tests/test\_cache.py::test\_cross\_scope\_invalidation} pins
the contract.

\paragraph{Performance footprint.} Cold global compute: $\sim$5–7 min
(one-time per machine; 35 models $\times$ 4 vars $\times$ 4 SSPs
NetCDF reads producing 1380 zarr entries totalling $\sim$3 GB). Warm
global + cold country: $\sim$1:45 (Greece), 1:42 (Sweden), 1:56
(Belgium). Warm cache (every artefact cached, 28 figures rendered):
$\sim$50 s, dominated by cartopy bias-map rendering. Further
optimisation (batched zarr reads, parallel figure render, bias-map
persistence) is a future performance sprint.

\paragraph{Decided.} 2026-05-02, in version 0.1.0.

% =============================================================================
\bibliographystyle{plainnat}
\bibliography{references}
% =============================================================================

\end{document}
